% Last change 2009-05-08
\input tex2page
\input plainsection
\input btxmac
\let\TZPtexlayout 0

\sidemargin 1.75 true in

\ifx\shipout\UNDEFINED

\let\PRIMsection\section
\def\section{\htmlpagebreak\PRIMsection}
\fi
\let\c\centerline
\let\n\noindent
\let\f\numberedfootnote

%\let\q\scm
\def\meta[#1]{{\em #1}}

\title{pregexp: \\
Portable Regular Expressions \\
for Scheme and Common Lisp}

\ifx\shipout\UnDeFiNeD
\c{\urlh{pregexp.tar.bz2}{Download}}
\fi

\c{\urlh{../index.html}{Dorai
Sitaram}}

\bigskip

\n \p{pregexp.scm} is a portable library for {\em
regular expressions} (aka {\em regexps} or {\em
regexes}) that runs in any Scheme
that complies with R4RS, R5RS or R6RS~\cite{r6rs}.  It
provides regular expressions
modeled on
Perl's~\cite{friedl:regex,pperl}, and includes such
powerful directives as numeric and nongreedy
quantifiers, capturing and non-capturing clustering,
POSIX character classes, selective case- and
space-insensitivity, backreferences, alternation,
backtrack pruning,
positive and negative lookahead and lookbehind, in
addition to the more basic directives familiar to all
regexp users.

To use, simply load the file \p{pregexp.scm} into
your Scheme.
(Alternatively, if your dialect allows it, you can
install \p{pregexp} as a {\em module} --- consult
the file \url{INSTALL} in the distribution.)

The file \p{pregexp.lisp} is the Common Lisp version
of this library.  To use it, load it into your Common Lisp.
(The descriptions and examples in this manual use Scheme
syntax, but their
translation to CL syntax is trivial.)

\htmlonly

\n{\bf Contents}

\tableofcontents
\endhtmlonly

\section{Introduction}

A {\em regexp} is a string that describes a pattern.  A
regexp matcher tries to {\em match} this pattern
against (a portion of) another string, which we
will call the {\em text string}.  The text string
is treated as raw text and not as a pattern.

Most of the characters in a regexp pattern are meant to
match occurrences of themselves in the text string.
Thus, the pattern \q{"abc"} matches a string that
contains the characters \p{a}, \p{b}, \p{c} in succession.

In the regexp pattern, some characters  act as {\em
metacharacters}, and some character sequences act as
{\em metasequences}.  That is, they specify something
other than their literal selves.  For example, in the
pattern \q{"a.c"}, the characters \p{a} and \p{c} do
stand for themselves but the {\em metacharacter} `\p{.}'
can match {\em any} character (other than
newline).  Therefore, the pattern \q{"a.c"}
matches an \p{a}, followed by {\em any} character,
followed by a \p{c}.

If we needed to match the character `\p{.}' itself,
we {\em escape} it, ie, precede it with a backslash
(\p{\}).  The character sequence \p{\.} is thus a {\em
metasequence}, since it doesn't match itself but rather
just `\p{.}'.  So, to match \p{a} followed by a literal
`\p{.}' followed by \p{c}, we use the regexp pattern
\q{"a\\.c"}.\f{The double backslash is an artifact of
Scheme strings, not the regexp pattern itself.  When we
want a literal backslash inside a Scheme string, we
must escape it so that it shows up in the string at
all. Scheme strings use backslash as the escape
character, so we end up with two backslashes --- one
Scheme-string backslash to escape the regexp backslash,
which then escapes the dot.  Another character that
would need escaping inside a Scheme string is
\ifx\newenvironment\PLAINTEX
`\p{"}'.
\else
`{\tt"}'.\fi
}
Another example of a metasequence is \p{\t}, which is a
readable way to represent the tab character.

We will call the string representation of a regexp the
{\em U-regexp}, where {\em U} can be taken to mean {\em
Unix-style} or {\em universal}, because this
notation for regexps is universally familiar.  Our
implementation uses an intermediate tree-like
representation called the {\em S-regexp}, where {\em S}
can stand for {\em Scheme}, {\em symbolic}, or
{\em s-expression}.  S-regexps are more verbose
and less readable than U-regexps, but they are much
easier for Scheme's recursive procedures to navigate.

%\section{Regexp procedures provided by \p{pregexp.scm}}
\section{Regexp procedures}

\p{pregexp.scm}
provides the procedures
\q{pregexp}, \q{pregexp-match-positions},
\q{pregexp-match}, \q{pregexp-split}, \q{pregexp-replace},
\q{pregexp-replace*}, and \q{pregexp-quote}.
All the identifiers introduced
by \p{pregexp.scm} have the prefix \q{pregexp}, so they
are unlikely to clash with other names in Scheme,
including those of any natively provided regexp
operators.

\subsection{\q{pregexp}}

The procedure \q{pregexp} takes
a U-regexp, which is a string, and returns
an S-regexp, which is a tree.

\q{
(pregexp "c.r")
=> (:sub (:or (:seq #\c :any #\r)))
}

\n There is rarely any need to look at the S-regexps
returned by \q{pregexp}.

\subsection{\q{pregexp-match-positions}}

The procedure \q{pregexp-match-positions} takes
a
regexp pattern and a text string, and returns a {\em
match} if the regexp {\em matches} (some part of) the text string.

The regexp may be either a U- or an S-regexp.
(\q{pregexp-match-positions} will internally compile a
U-regexp to an S-regexp before proceeding with the
matching.  If you find yourself calling
\q{pregexp-match-positions} repeatedly with the same
U-regexp, it may be advisable to explicitly convert the
latter into an S-regexp once beforehand, using
\q{pregexp}, to save needless recompilation.)

\q{pregexp-match-positions} returns \q{#f} if the regexp did not
match the string; and a list of {\em index pairs} if it
did match.  Eg,

\q{
(pregexp-match-positions "brain" "bird")
=> #f

(pregexp-match-positions "needle" "hay needle stack")
=> ((4 . 10))
}

\n In the second example, the integers 4 and 10 identify
the substring that was matched. 4 is the starting
(inclusive) index and 10 the ending (exclusive) index of
the matching substring.

\q{
(substring "hay needle stack" 4 10)
=> "needle"
}

Here, \q{pregexp-match-positions}'s return list contains only
one index pair, and that pair represents the entire
substring matched by the regexp.  When we discuss
{\em subpatterns} later, we will see how a single match
operation can yield a list of {\em submatches}.

\q{pregexp-match-positions} takes optional third
and fourth arguments that specify the indices of
the text string within which the matching should
take place.

\q{
(pregexp-match-positions "needle"
  "his hay needle stack -- my hay needle stack -- her hay needle stack"
  24 43)
=> ((31 . 37))
}

\n Note that the returned indices are still reckoned
relative to the full text string.

\subsection{\q{pregexp-match}}

The procedure \q{pregexp-match} is called
like \q{pregexp-match-positions}
but instead of returning index pairs it returns the
matching substrings:

\q{
(pregexp-match "brain" "bird")
=> #f

(pregexp-match "needle" "hay needle stack")
=> ("needle")
}

\n \q{pregexp-match} also takes optional third and
fourth arguments, with the same meaning as does
\q{pregexp-match-positions}.

\subsection{\q{pregexp-split}}

The procedure \q{pregexp-split} takes
two arguments, a
regexp pattern and a text string, and returns a list of
substrings of the text string, where the pattern identifies the
delimiter separating the substrings.

\q{
(pregexp-split ":" "/bin:/usr/bin:/usr/bin/X11:/usr/local/bin")
=> ("/bin" "/usr/bin" "/usr/bin/X11" "/usr/local/bin")

(pregexp-split " " "pea soup")
=> ("pea" "soup")
}

\n If the first argument can match an empty string, then
the list of all the single-character substrings is returned.

\q{
(pregexp-split "" "smithereens")
=> ("s" "m" "i" "t" "h" "e" "r" "e" "e" "n" "s")
}

\n To identify one-or-more spaces as the delimiter,
take care to use the regexp \q{" +"}, not \q{" *"}.

\q{
(pregexp-split " +" "split pea     soup")
=> ("split" "pea" "soup")

(pregexp-split " *" "split pea     soup")
=> ("s" "p" "l" "i" "t" "p" "e" "a" "s" "o" "u" "p")
}

\subsection{\q{pregexp-replace}}

The procedure \q{pregexp-replace} replaces
the
matched portion of the text string by another
string.  The first argument is the pattern,
the second the text string, and the third
is the {\em insert string} (string to be inserted).

\q{
(pregexp-replace "te" "liberte" "ty")
=> "liberty"
}

\n If the pattern doesn't occur in the text
string, the returned string is identical (\q{eq?})
to the text string.

\subsection{\q{pregexp-replace*}}

The procedure \q{pregexp-replace*} replaces
{\em all}
matches in the text string by the insert
string:

\q{
(pregexp-replace* "te" "liberte egalite fraternite" "ty")
=> "liberty egality fratyrnity"
}

\n As with \q{pregexp-replace}, if the pattern doesn't
occur in the text string, the returned string is
identical (\q{eq?}) to the text string.

\subsection{\q{pregexp-quote}}

The procedure \q{pregexp-quote} takes
an arbitrary string and returns a U-regexp
(string) that precisely represents it.  In particular,
characters in the input string that could serve as
regexp metacharacters are escaped with a
backslash, so that they safely match only themselves.

\q{
(pregexp-quote "cons")
=> "cons"

(pregexp-quote "list?")
=> "list\\?"
}

\q{pregexp-quote} is useful when building a composite
regexp from a mix of regexp strings and verbatim strings.

\section{The regexp pattern language}

Here is a complete description of the regexp pattern
language recognized by the \q{pregexp} procedures.

\subsection{Basic assertions}

The {\em assertions} \p{^} and \p{$} identify the
beginning and the end of the text string respectively.
They ensure that their adjoining regexps match at
one or other end of the text string.
Examples:

\q{
(pregexp-match-positions "^contact" "first contact")
=> #f
}

\n The regexp fails to match because \p{contact} does not
occur at the beginning of the text string.

\q{
(pregexp-match-positions "laugh$" "laugh laugh laugh laugh")
=> ((18 . 23))
}

\n The regexp matches the {\em last} \p{laugh}.

The metasequence \p{\b} asserts that
a {\em word boundary} exists.

\q{
(pregexp-match-positions "yack\\b" "yackety yack")
=> ((8 . 12))
}

\n The \p{yack} in \p{yackety} doesn't end at a word
boundary so it isn't matched.  The second \p{yack} does
and is.

The metasequence \p{\B} has the opposite effect
to \p{\b}.   It asserts that a word boundary
does not exist.

\q{
(pregexp-match-positions "an\\B" "an analysis")
=> ((3 . 5))
}

\n The \p{an} that doesn't end in a word boundary
is matched.

\subsection{Characters and character classes}

Typically a character in the regexp matches the same
character in the text string.  Sometimes it is
necessary or convenient to use a regexp
metasequence to refer to a single character.
Thus, metasequences \p{\n}, \p{\r}, \p{\t}, and \p{\.}
match the newline, return, tab and period characters
respectively.

The {\em metacharacter} period (\p{.}) matches
{\em any} character other than newline.

\q{
(pregexp-match "p.t" "pet")
=> ("pet")
}

\n It also matches \p{pat}, \p{pit}, \p{pot}, \p{put},
and \p{p8t} but not \p{peat} or \p{pfffft}.

A {\em character class} matches any one character from
a set of characters.  A typical format for this
is the {\em bracketed character class} \p{[}...\p{]},
which matches any one character from the non-empty sequence
of characters enclosed within the brackets.\f{Requiring
a bracketed character class to be non-empty is not a limitation,
since an
empty character class
can be more easily represented by an empty string.}
Thus \q{"p[aeiou]t"} matches \p{pat}, \p{pet}, \p{pit},
\p{pot}, \p{put} and nothing else.

Inside the brackets, a hyphen (\p{-}) between two
characters specifies the ascii range between the characters.
Eg, \q{"ta[b-dgn-p]"} matches \p{tab}, \p{tac}, \p{tad}, {\em and}
\p{tag}, {\em and} \p{tan}, \p{tao}, \p{tap}.

An initial caret (\p{^}) after the left bracket inverts
the set specified by the rest of the contents, ie, it
specifies the set of characters {\em other than} those
identified in the brackets.  Eg, \q{"do[^g]"} matches
all three-character sequences starting with \p{do}
except \p{dog}.

Note that the metacharacter \p{^} inside brackets means
something quite different from what it means outside.
Most other metacharacters (\p{.}, \p{*}, \p{+}, \p{?},
etc) cease to be metacharacters when inside brackets,
although you may still escape them for peace of
mind.  \p{-} is a metacharacter only when it's
inside brackets, and neither the first nor the last character.

Bracketed character classes cannot contain other
bracketed character classes (although they contain
certain other types of character classes --- see
below).  Thus a left bracket (\p{[})
inside a bracketed character class doesn't have to be a
metacharacter; it can stand for itself.  Eg,
\q{"[a[b]"} matches \p{a}, \p{[}, and \p{b}.

Furthermore, since empty bracketed character classes
are disallowed, a right bracket (\p{]}) immediately occurring
after the opening left bracket
also doesn't need to be a metacharacter.  Eg,
\q{"[]ab]"} matches \p{]}, \p{a}, and \p{b}.

\subsubsection{Some frequently used character
classes}

Some standard character classes can be conveniently
represented as metasequences instead of as explicit
bracketed expressions.  \p{\d} matches a  digit
(\p{[0-9]}); \p{\s} matches a whitespace character; and
\p{\w} matches a character that could be part of a
``word''.\f{Following regexp custom, we identify
``word'' characters as
\ifx\newenvironment\PLAINTEX
\p{[A-Za-z0-9_]}\else
{\tt[A-Za-z0-9\_]}\fi
, although these
are too restrictive for what a Schemer might consider a
``word''.}

The upper-case versions of these metasequences stand
for the inversions of the corresponding character
classes.  Thus \p{\D} matches a non-digit, \p{\S} a
non-whitespace character, and \p{\W} a
non-``word'' character.

Remember to include a double backslash when putting
these metasequences in a Scheme string:

\q{
(pregexp-match "\\d\\d"
  "0 dear, 1 have 2 read catch 22 before 9")
=> ("22")
}

These character classes can be used inside
a bracketed expression.  Eg,
\q{"[a-z\\d]"} matches a lower-case letter
or a digit.

\subsubsection{POSIX character classes}

A {\em POSIX character class} is a special metasequence
of the form \p{[:}...\p{:]} that can be used only
inside a bracketed expression.  The POSIX classes
supported are

\medskip

\halign{\indent #\hfil & \quad #\hfil \cr
\p{[:alnum:]}  & letters and digits \cr
\p{[:alpha:]}  & letters  \cr
\p{[:algor:]}  & the letters \p{c}, \p{h}, \p{a} and \p{d} \cr
\p{[:ascii:]}  & 7-bit ascii characters \cr
\p{[:blank:]}  & widthful whitespace, ie, space and tab \cr
\p{[:cntrl:]}  & ``control'' characters, viz, those with code \p{<} 32 \cr
\p{[:digit:]}  & digits, same as \p{\d} \cr
\p{[:graph:]}  & characters that use ink \cr
\p{[:lower:]}  & lower-case letters \cr
\p{[:print:]}  & ink-users plus widthful whitespace \cr
\p{[:space:]}  & whitespace, same as \p{\s} \cr
\p{[:upper:]}  & upper-case letters \cr
\p{[:word:]}   & letters, digits, and underscore, same as \p{\w} \cr
\p{[:xdigit:]} & hex digits \cr
}

\medskip

\n For example, the regexp  \q{"[[:alpha:]_]"}
matches a letter or underscore.

\q{
(pregexp-match "[[:alpha:]_]" "--x--")
=> ("x")

(pregexp-match "[[:alpha:]_]" "--_--")
=> ("_")

(pregexp-match "[[:alpha:]_]" "--:--")
=> #f
}

The POSIX class notation is valid {\em only} inside a
bracketed expression.  For instance, \p{[:alpha:]},
when not inside a bracketed expression, will {\em not}
be read as the letter class.
Rather it is (from previous principles) the character
class containing the characters \p{:}, \p{a}, \p{l},
\p{p}, \p{h}.

\q{
(pregexp-match "[:alpha:]" "--a--")
=> ("a")

(pregexp-match "[:alpha:]" "--_--")
=> #f
}

By placing a caret (\p{^}) immediately after
\p{[:}, you get the inversion of that POSIX
character class.  Thus, \p{[:^alpha]}
is the class containing all characters
except the letters.

\subsection{Quantifiers}

The {\em quantifiers} \p{*}, \p{+}, and
\p{?} match respectively: zero or more, one or more,
and zero or one instances of the preceding subpattern.

\q{
(pregexp-match-positions "c[ad]*r" "cadaddadddr")
=> ((0 . 11))
(pregexp-match-positions "c[ad]*r" "cr")
=> ((0 . 2))

(pregexp-match-positions "c[ad]+r" "cadaddadddr")
=> ((0 . 11))
(pregexp-match-positions "c[ad]+r" "cr")
=> #f

(pregexp-match-positions "c[ad]?r" "cadaddadddr")
=> #f
(pregexp-match-positions "c[ad]?r" "cr")
=> ((0 . 2))
(pregexp-match-positions "c[ad]?r" "car")
=> ((0 . 3))
}

\subsubsection{Numeric quantifiers}

You can use braces to specify much finer-tuned
quantification than is possible with \p{*}, \p{+}, \p{?}.

The quantifier \p{{m}} matches {\em exactly} \p{m}
instances of the preceding {\em subpattern}.  \p{m}
must be a nonnegative integer.

The quantifier \p{{m,n}} matches at least \p{m}
and at most \p{n} instances.  \p{m} and
\p{n} are nonnegative integers with \p{m <=
n}.  You may omit either or both numbers, in which case
\p{m} defaults to 0 and \p{n} to
infinity.

It is evident that \p{+} and \p{?} are abbreviations
for \p{{1,}} and \p{{0,1}} respectively.
\p{*} abbreviates \p{{,}}, which is the same
as \p{{0,}}.

\q{
(pregexp-match "[aeiou]{3}" "vacuous")
=> ("uou")

(pregexp-match "[aeiou]{3}" "evolve")
=> #f

(pregexp-match "[aeiou]{2,3}" "evolve")
=> #f

(pregexp-match "[aeiou]{2,3}" "zeugma")
=> ("eu")
}

\subsubsection{Non-greedy quantifiers}

The quantifiers described above are {\em greedy}, ie,
they match the  maximal number of instances that would
still lead to an overall match for the full pattern.

\q{
(pregexp-match "<.*>" "<tag1> <tag2> <tag3>")
=> ("<tag1> <tag2> <tag3>")
}

To make these quantifiers {\em non-greedy}, append
a \p{?} to them.  Non-greedy quantifiers match
the minimal number of instances needed to ensure an
overall match.

\q{
(pregexp-match "<.*?>" "<tag1> <tag2> <tag3>")
=> ("<tag1>")
}

The non-greedy quantifiers are respectively:
\p{*?}, \p{+?}, \p{??}, \p{{m}?}, \p{{m,n}?}.
Note the two uses of the metacharacter \p{?}.

\subsection{Clusters}

{\em Clustering}, ie, enclosure within parens
\p{(}...\p{)}, identifies the enclosed {\em subpattern}
as a single entity.  It causes the matcher to {\em capture}
the {\em submatch}, or the portion of the string
matching the subpattern, in addition to the
overall match.

\q{
(pregexp-match "([a-z]+) ([0-9]+), ([0-9]+)" "jan 1, 1970")
=> ("jan 1, 1970" "jan" "1" "1970")
}

Clustering also causes a following quantifier to treat
the entire enclosed subpattern as an entity.

\q{
(pregexp-match "(poo )*" "poo poo platter")
=> ("poo poo " "poo ")
}

The number of submatches returned is always equal
to the number of subpatterns specified in the
regexp, even if a particular subpattern happens
to match more than one substring or no substring
at all.

\q{
(pregexp-match "([a-z ]+;)*" "lather; rinse; repeat;")
=> ("lather; rinse; repeat;" " repeat;")
}

\n Here the \p{*}-quantified subpattern matches three
times, but it is the last submatch that is returned.

It is also possible for a quantified subpattern to
fail to match, even if the overall pattern matches.
In such cases, the failing submatch is represented
by \q{#f}.

\q{
(define date-re
  ;match `month year' or `month day, year'.
  ;subpattern matches day, if present
  (pregexp "([a-z]+) +([0-9]+,)? *([0-9]+)"))

(pregexp-match date-re "jan 1, 1970")
=> ("jan 1, 1970" "jan" "1," "1970")

(pregexp-match date-re "jan 1970")
=> ("jan 1970" "jan" #f "1970")
}

\subsubsection{Backreferences}

Submatches can be used in the insert string argument of
the procedures \q{pregexp-replace} and
\q{pregexp-replace*}.  The insert string can use \p{\n}
as a {\em backreference} to refer back to the {\em n}th
submatch, ie, the substring that matched the {\em n}th
subpattern.   \p{\0} refers to the entire match,
and it can also be specified as \p{\&}.

\q{
(pregexp-replace "_(.+?)_"
  "the _nina_, the _pinta_, and the _santa maria_"
  "*\\1*")
=> "the *nina*, the _pinta_, and the _santa maria_"

(pregexp-replace* "_(.+?)_"
  "the _nina_, the _pinta_, and the _santa maria_"
  "*\\1*")
=> "the *nina*, the *pinta*, and the *santa maria*"

;recall: \S stands for non-whitespace character

(pregexp-replace "(\\S+) (\\S+) (\\S+)"
  "eat to live"
  "\\3 \\2 \\1")
=> "live to eat"
}

Use \p{\\} in the insert string to specify a literal
backslash.  Also, \p{\$} stands for an empty string,
and is useful for separating a backreference \p{\n}
from an immediately following number.

Backreferences can also be used within the regexp
pattern to refer back to an already matched subpattern
in the pattern.  \p{\n} stands for an exact repeat
of the {\em n}th submatch.\f{\ifx\newenvironment\PLAINTEX
\p{\0}\else
{\tt\\0}\fi
, which is useful in
an insert string, makes no  sense within the regexp
pattern, because the entire regexp has not matched yet
that you could refer back to it.}

\q{
(pregexp-match "([a-z]+) and \\1"
  "billions and billions")
=> ("billions and billions" "billions")
}

\n Note that the backreference is not simply a repeat
of the previous subpattern.  Rather it is a repeat of
{\em the particular  substring already matched by the
subpattern}.

In the above example, the backreference can only match
\p{billions}.  It will not match \p{millions}, even
though the subpattern it harks back to --- \p{([a-z]+)}
---  would have had no problem doing so:

\q{
(pregexp-match "([a-z]+) and \\1"
  "billions and millions")
=> #f
}

The following corrects doubled words:

\q{
(pregexp-replace* "(\\S+) \\1"
  "now is the the time for all good men to to come to the aid of of the party"
  "\\1")
=> "now is the time for all good men to come to the aid of the party"
}

The following marks all immediately repeating patterns
in a number string:

\q{
(pregexp-replace* "(\\d+)\\1"
  "123340983242432420980980234"
  "{\\1,\\1}")
=> "12{3,3}40983{24,24}3242{098,098}0234"
}

\subsubsection{Non-capturing clusters}

It is often required to specify a cluster
(typically for quantification) but without triggering
the capture of submatch information.  Such
clusters are called {\em non-capturing}.  In such cases,
use \p{(?:} instead of \p{(} as the cluster opener.  In
the following example, the  non-capturing cluster
eliminates the ``directory'' portion of a given
pathname, and the capturing cluster  identifies the
basename.

\q{
(pregexp-match "^(?:[a-z]*/)*([a-z]+)$"
  "/usr/local/bin/mzscheme")
=> ("/usr/local/bin/mzscheme" "mzscheme")
}

\subsubsection{Cloisters}

The location between the \p{?} and the \p{:} of a
non-capturing cluster is called a {\em cloister}.\f{A
useful, if terminally cute, coinage from the abbots of
Perl~\cite{pperl}.}  You can put {\em modifiers}
there that will cause the enclustered subpattern to be
treated specially.  The modifier \p{i} causes the
subpattern to match {\em case-insensitively}:

\q{
(pregexp-match "(?i:hearth)" "HeartH")
=> ("HeartH")
}

The modifier \p{x} causes the subpattern to match
{\em space-insensitively}, ie, spaces and
comments within the
subpattern are ignored.  Comments are introduced
as usual with a semicolon (\p{;}) and extend till
the end of the line.  If you need
to include a literal space or semicolon in
a space-insensitized subpattern, escape it
with a backslash.

\q{
(pregexp-match "(?x: a   lot)" "alot")
=> ("alot")

(pregexp-match "(?x: a  \\  lot)" "a lot")
=> ("a lot")

(pregexp-match "(?x:
   a \\ man  \\; \\   ; ignore
   a \\ plan \\; \\   ; me
   a \\ canal         ; completely
   )"
 "a man; a plan; a canal")
=> ("a man; a plan; a canal")
}

\n The global variable \q{*pregexp-comment-char*}
contains the comment character (\q{#\;}).
For Perl-like comments,

\q{
(set! *pregexp-comment-char* #\#)
}

You can put more than one modifier in the cloister.

\q{
(pregexp-match "(?ix:
   a \\ man  \\; \\   ; ignore
   a \\ plan \\; \\   ; me
   a \\ canal         ; completely
   )"
 "A Man; a Plan; a Canal")
=> ("A Man; a Plan; a Canal")
}

A minus sign before a modifier inverts its meaning.
Thus, you can use \p{-i} and \p{-x} in a {\em
subcluster} to overturn the insensitivities caused by an
enclosing cluster.

\q{
(pregexp-match "(?i:the (?-i:TeX)book)"
  "The TeXbook")
=> ("The TeXbook")
}

\n This regexp will allow any casing for \p{the}
and \p{book} but insists that \p{TeX} not be
differently cased.

\subsection{Alternation}
\label{alternation}

You can specify a list of {\em alternate}
subpatterns by separating them by \p/|/.   The \p/|/
separates subpatterns in the nearest enclosing cluster
(or in the entire pattern string if there are no
enclosing parens).

\q@
(pregexp-match "f(ee|i|o|um)" "a small, final fee")
=> ("fi" "i")

(pregexp-replace* "([yi])s(e[sdr]?|ing|ation)"
   "it is energising to analyse an organisation
   pulsing with noisy organisms"
   "\\1z\\2")
=> "it is energizing to analyze an organization
   pulsing with noisy organisms"
@

Note again that if you wish
to use clustering merely to specify a list of alternate
subpatterns but do not want the submatch, use \p{(?:}
instead of \p{(}.

\q@
(pregexp-match "f(?:ee|i|o|um)" "fun for all")
=> ("fo")
@

An important thing to note about alternation is that
the leftmost matching alternate is picked regardless of
its length.  Thus, if one of the alternates is a prefix
of a later alternate, the latter may not have
a chance to match.

\q@
(pregexp-match "call|call-with-current-continuation"
  "call-with-current-continuation")
=> ("call")
@

To allow the longer alternate to have a shot at
matching, place it before the shorter one:

\q@
(pregexp-match "call-with-current-continuation|call"
  "call-with-current-continuation")
=> ("call-with-current-continuation")
@

In any case, an overall match for the entire regexp is
always preferred to an overall nonmatch.  In the
following, the longer alternate still wins, because its
preferred shorter prefix fails to yield an overall
match.

\q@
(pregexp-match "(?:call|call-with-current-continuation) constrained"
  "call-with-current-continuation constrained")
=> ("call-with-current-continuation constrained")
@

\subsection{Backtracking}

We've already seen that greedy quantifiers match
the maximal number of times, but the overriding priority
is that the overall match succeed.  Consider

\q{
(pregexp-match "a*a" "aaaa")
}

\n The regexp consists of two subregexps,
\p{a*} followed by \p{a}.
The subregexp \p{a*} cannot be allowed to match
all four \p{a}'s in the text string \p{"aaaa"}, even though
\p{*} is a greedy quantifier.  It may match only the first
three, leaving the last one for the second subregexp.
This ensures that the full regexp matches successfully.

The regexp matcher accomplishes this via a process
called {\em backtracking}.  The matcher
tentatively allows the greedy quantifier
to match all four \p{a}'s, but then when it becomes
clear that the overall match is in jeopardy, it
{\em backtracks} to a less greedy match of {\em
three} \p{a}'s.  If even this fails, as in the
call

\q{
(pregexp-match "a*aa" "aaaa")
}

\n the matcher backtracks even further.  Overall
failure is conceded only when all possible backtracking
has been tried with no success.

Backtracking is not restricted to greedy quantifiers.
Nongreedy quantifiers match as few instances as
possible, and progressively backtrack to more and more
instances in order to attain an overall match.  There
is backtracking in alternation too, as the more
rightward alternates are tried when locally successful
leftward ones fail to yield an overall match.

\subsubsection{Disabling backtracking}

Sometimes it is efficient to disable backtracking.  For
example, we may wish  to  {\em commit} to a choice, or
we know that trying alternatives is fruitless.  A
nonbacktracking regexp is enclosed in \p{(?>}...\p{)}.

\q{
(pregexp-match "(?>a+)." "aaaa")
=> #f
}

In this call, the subregexp \p{?>a+} greedily matches
all four \p{a}'s, and is denied the opportunity to
backpedal.  So the overall match is denied.  The effect
of the regexp is therefore to match one or more \p{a}'s
followed by something that is definitely non-\p{a}.

\subsection{Looking ahead and behind}

You can have assertions in your pattern that look {\em
ahead} or {\em behind} to ensure that a subpattern does
or does not occur.   These ``look around'' assertions are
specified by putting the subpattern checked for in a
cluster whose leading characters are: \p{?=} (for positive
lookahead), \p{?!} (negative lookahead), \p{?<=}
(positive lookbehind), \p{?<!} (negative lookbehind).
Note that the subpattern in the assertion  does not
generate a match in the final result.  It merely allows
or disallows the rest of the match.

\subsubsection{Lookahead}

Positive lookahead (\p{?=}) peeks ahead to ensure that
its subpattern {\em could} match.

\q{
(pregexp-match-positions "grey(?=hound)"
  "i left my grey socks at the greyhound")
=> ((28 . 32))
}

\n The regexp \q{"grey(?=hound)"} matches \p{grey}, but
{\em only} if it is followed by \p{hound}.  Thus, the first
\p{grey} in the text string is not matched.

Negative lookahead (\p{?!}) peeks ahead
to ensure that its subpattern could not possibly match.

\q{
(pregexp-match-positions "grey(?!hound)"
  "the gray greyhound ate the grey socks")
=> ((27 . 31))
}

\n The regexp \q{"grey(?!hound)"} matches \p{grey}, but
only if it is {\em not} followed by \p{hound}.  Thus
the \p{grey} just before \p{socks} is matched.

\subsubsection{Lookbehind}

Positive lookbehind (\p{?<=}) checks that its subpattern {\em could} match
immediately to the left of the current position in
the text string.

\q{
(pregexp-match-positions "(?<=grey)hound"
  "the hound in the picture is not a greyhound")
=> ((38 . 43))
}

\n The regexp \p{(?<=grey)hound} matches \p{hound}, but only if it is
preceded by \p{grey}.

Negative lookbehind
(\p{?<!}) checks that its subpattern
could not possibly match immediately to the left.

\q{
(pregexp-match-positions "(?<!grey)hound"
  "the greyhound in the picture is not a hound")
=> ((38 . 43))
}

\n The regexp \p{(?<!grey)hound} matches \p{hound}, but only if
it is {\em not} preceded by \p{grey}.

Lookaheads and lookbehinds can be convenient when they
are not confusing.

\section{An extended example}

Here's an extended example from
Friedl~\cite[p 189]{friedl:regex}
that covers many of the features described
above.  The problem is to fashion a regexp that will
match any and only IP addresses or {\em dotted
quads}, ie, four numbers separated by three dots, with
each number between 0 and 255.  We will use the
commenting mechanism to build the final regexp with
clarity.  First, a subregexp \q{n0-255} that matches 0
through 255.

\q{
(define n0-255
  "(?x:
  \\d          ;  0 through   9
  | \\d\\d     ; 00 through  99
  | [01]\\d\\d ;000 through 199
  | 2[0-4]\\d  ;200 through 249
  | 25[0-5]    ;250 through 255
  )")
}

\n The first two alternates simply get all single- and
double-digit numbers.  Since 0-padding is allowed, we
need to match both 1 and 01.  We need to be careful
when getting 3-digit numbers, since numbers above 255
must be excluded.  So we fashion alternates to get 000
through 199, then 200 through 249, and finally 250
through 255.\f{Note that
\ifx\newenvironment\PLAINTEX
\q{n0-255}
\else
\var{n0-255}
\fi
lists prefixes as
preferred alternates, something we cautioned against in
sec \ref{alternation}.  However, since we intend
to anchor this subregexp explicitly to force an overall
match, the order of the alternates does not matter.}

An IP-address is a string that consists of
four \q{n0-255}s with three dots separating
them.

\q{
(define ip-re1
  (string-append
    "^"        ;nothing before
    n0-255     ;the first n0-255,
    "(?x:"     ;then the subpattern of
    "\\."      ;a dot followed by
    n0-255     ;an n0-255,
    ")"        ;which is
    "{3}"      ;repeated exactly 3 times
    "$"        ;with nothing following
    ))
}

\n Let's try it out.

\q{
(pregexp-match ip-re1
  "1.2.3.4")
=> ("1.2.3.4")

(pregexp-match ip-re1
  "55.155.255.265")
=> #f
}

\n which is fine, except that we also
have

\q{
(pregexp-match ip-re1
  "0.00.000.00")
=> ("0.00.000.00")
}

\n All-zero sequences are not valid IP addresses!
Lookahead to the rescue.  Before starting to match
\q{ip-re1}, we look ahead to ensure we don't have all
zeros.  We could use positive lookahead
to ensure there {\em is} a digit other than
zero.

\q{
(define ip-re
  (string-append
    "(?=.*[1-9])" ;ensure there's a non-0 digit
    ip-re1))
}

\n Or we could use negative lookahead to
ensure that what's ahead isn't composed
of {\em only} zeros and dots.

\q{
(define ip-re
  (string-append
    "(?![0.]*$)" ;not just zeros and dots
                 ;(note: dot is not metachar inside [])
    ip-re1))
}

\n The regexp \q{ip-re} will match
all and only valid IP addresses.

\q{
(pregexp-match ip-re
  "1.2.3.4")
=> ("1.2.3.4")

(pregexp-match ip-re
  "0.0.0.0")
=> #f
}

\section{References}

%\nocite{sre,nregex,kleene:regexp}

\thebibliography{99}
\bibitem{friedl:regex} Jeffrey E. F. Friedl,
\urlh{http://regex.info}{Mastering Regular Expressions}, 2/e, O'Reilly,
2002.

\bibitem{r6rs} Michael Sperber, R. Kent Dybvig, Matthew Flatt, and Anton
van Straaten (eds.), \urlh{http://www.r6rs.org}{Revised$^6$ Report on the Algorithmic Language
Scheme (``R6RS'')}, 2007.

\bibitem{pperl} Larry Wall, Tom Christensen, and Jon Orwant,
\urlh{http://oreilly.com/catalog/9780596000271}{Programming Perl},
3/e, O'Reilly, 2000.
\endthebibliography

\bye
